---
title: QPAD bug explanation
author: Elly Knight
date: May 7, 2025
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load libraries
library(raster)
library(sp)
library(suntools)
library(lutz)
library(suncalc)
library(tidyverse)

#load time offset zone raster
rtz <- raster("utcoffset.tif")

#get example data
dat <- read.csv("TSSRExample.csv")

#format to dtm
date <- dat$datetime
dtm <- strptime(date, format="%Y-%m-%d %H:%M:%S", tz="America/Edmonton")

#get day and hour
day <- as.integer(dtm$yday)
hour <- as.numeric(round(dtm$hour + dtm$min/60, 2))

#get coords
lon <- dat$lon
lat <- dat$lat

#project for extraction
xy <- data.frame(x=lon, y=lat)
xy$x[is.na(xy$x)] <- mean(xy$x, na.rm=TRUE)
xy$y[is.na(xy$y)] <- mean(xy$y, na.rm=TRUE)
coordinates(xy) <- ~ x + y
proj4string(xy) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
xy <- invisible(spTransform(xy, crs(rtz)))
```

## Background

In QPAD, the set of removal models for availability for detection includes a term for time of day, measures as time since sunrise (TSSR). For each species, a set of models are fit and the best-fitting model is selected as the model with the lowest BIC.

The coefficients from that best-fitting model are then used to calculate the mean cue rate for a species as part of the QPAD offset calculation. The functions within the `qpad-offsets` repository and the `wildrtrax` R package that prepare the covariates for cue rate calculation calculate TSSR for the mountain time zone and then use the coordinates of the survey to extract a time zone value for that location from a raster and use that time zone value to adjust the TSSR. We use this raster-based correction approach because it is much more efficient for large datasets than other available time zone lookup functions (see 'Comparison to alternate approach' below)

```{r, echo=FALSE}

rtz.df <- rtz |> 
  raster::aggregate(10) |> 
  raster::projectRaster(crs=4326) |> 
  as.data.frame(xy=TRUE) |> 
  mutate(utcoffset = round(utcoffset)) |> 
  dplyr::filter(!utcoffset %in% c("NaN", -3),
                !is.na(utcoffset))

ggplot() +
  geom_raster(data=rtz.df, aes(x=x, y=y, fill=utcoffset)) +
  geom_point(data=dat, aes(x=lon, y=lat)) +
  ggtitle("Time zone raster with example survey locations") +
  scale_fill_viridis_c(name="Time zone\noffset") +
  theme_classic() +
  xlab("Longitude") +
  ylab("Latitude") 

```

## Code bug

In the original code, the adjustment was made by subtracting sunrise from the survey time and adding the time zone offset.

```{r}

# get time zone offset
ltz <- raster::extract(rtz, xy) + 7

# get sunrise in mountain time
sr <- sunriset(cbind("X"=lon, "Y"=lat),
               as.POSIXct(dtm, tz="America/Edmonton"),
               direction="sunrise", POSIXct.out=FALSE) * 24

# calculate TSSR
TSSR_bug <- hour - sr$day_frac + ltz
```

However, the addition of the time zone offset doubles the time zone difference relative to mountain time zone instead of zeroing it. We can detect this bug in the TSSR as a linear relationship between TSSR and longitude.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
tssr.df <- data.frame(ltz=ltz,
                      hour=hour,
                      sr=sr$day_frac,
                      lon=lon,
                      TSSR1=TSSR_bug)

ggplot(tssr.df) +
  geom_point(aes(x=lon, y=TSSR1)) +
  theme_classic() +
  xlab("Longitude") +
  ylab("Hours since sunrise")

```

## Bug Correction

If we instead subtract the time zone offset, we see the relationship is nullified.
```{r}
# calculate TSSR
TSSR_fix <- hour - sr$day_frac - ltz
```
```{r, echo=FALSE, warning=FALSE}
tssr.df <- data.frame(ltz=ltz,
                      hour=hour,
                      sr=sr$day_frac,
                      lon=lon,
                      TSSR=TSSR_fix)

ggplot(tssr.df) +
  geom_point(aes(x=lon, y=TSSR)) +
  theme_classic() +
  xlab("Longitude") +
  ylab("Hours since sunrise")

```

## Comparison to alternate approach

We can double check that subtraction of the time zone offset gives us the correct TSSR by comparing to a slower alternate approach. Here, we use the `lutz` package to look up timezone using the coordinates and use that timezone in the sunrise calculation directly.
```{r}
#look up timezones
dat$tz <- tz_lookup_coords(lat, lon, method="accurate")

#calculate sunrise for each timezone
tzs <- unique(dat$tz)
out <- data.frame()
for(i in 1:length(tzs)){

    #get just the data for that timezone
    dat.i <- dplyr::filter(dat, tz==tzs[i]) |>
        mutate(datetime = ymd_hms(datetime, tz=tzs[i]),
               date = as.Date(datetime, tz=tzs[i]))

    #get sunrise time
    dat.i$sr2 <- getSunlightTimes(data=dat.i, keep="sunrise", tz=tzs[i])$sunrise
    
    #fix America/Regina
    if(tzs[i]=="America/Regina"){dat.i$sr2 <- dat.i$sr2 + 3600}

    #output
    out <- rbind(out, dat.i)
}

#calculate tssr
out$TSSR_alternate <- as.numeric(difftime(out$datetime, out$sr), units="hours")
```
Let's compare `TSSR_bug` and `TSSR_fix` to the alernate approach to double check.
```{r, echo=FALSE}
check <- dat |> 
  mutate(TSSR_bug = TSSR_bug,
         TSSR_fix = TSSR_fix,
         ltz = ltz - 7) |> 
  left_join(out |> 
              dplyr::select(-datetime)) |> 
  pivot_longer(TSSR_bug:TSSR_fix, names_to="approach", values_to="TSSR") |> 
  mutate(diff = TSSR_alternate - TSSR)

ggplot(check |> dplyr::filter(!is.na(TSSR))) +
  geom_point(aes(x=TSSR_alternate, y=TSSR, colour=ltz)) +
  facet_wrap(~approach) +
  scale_colour_viridis_c(name="Time zone\noffset") +
  theme_classic() +
  xlab("Hours since sunrise from alternate approach") +
  ylab("Hours since sunrise from QPAD approach") +
  theme(legend.position="bottom")
```
Looks good, onward!
