---
title: Adjusting offsets
output: pdf_document
---

In QPAD, we have the expected value of the counts ($Y$) defined as $E[Y]=DApq$.
We use the following settings for density ($D$) and the detectability correction ($Apq$).
The _old_ (biased due to TZ offset issue) and _new_ (fixed) corrections
are used here as follows:

```{r}
D <- 1
Apq_old <- 0.94
Apq_new <- 1.12
```

These numbers were taken from the actual offsets (mean of the means) from the BAM data.

Let's assume that the _new_ correction is right, so we simulate under:

```{r}
n <- 1000
Y <- rpois(n, lambda = D * Apq_new)
```

Now we estimate density as if our detectability offset estimate would be correct.
We can see that the estimate of $D$ is correct, as expected:

```{r}
off_new <- log(rep(Apq_new, n))
m1 <- glm(Y ~ 1 + offset(off_new), family=poisson)
(D_new <- exp(coef(m1)))
```

Now what happens if we use the biased offset estimate? $D$ is overestimated:

```{r}
off_old <- log(rep(Apq_old, n))
m2 <- glm(Y ~ 1 + offset(off_old), family=poisson)
(D_old <- exp(coef(m2)))
```

The question now: how can we correct the _old_ density estimate using the
_old_ and _new_ offsets without having to re-estimate density using the _new_ offsets.

$Y=D_{old}C_{old}$ and $Y=D_{new}C_{new}$, therefore
$D_{new} = D_{old}C_{old}/C_{new}$.
On the log scale: $log(D_{new}) = log(D_{old})+log(C_{old})-log(C_{new})$.

```{r}
adj <-  log(Apq_old) - log(Apq_new)
exp(log(D_old) + adj)
```

Which is the same as:

```{r}
alpha <- exp(adj)
D_old * alpha
```

Applying the adjustment to tabular population summaries:

- calculate the exponentiated adjustment value as $\alpha=C_{old}/C_{new}$
- apply the $\alpha$ value on density (not estimated population size $N$) so that $D_{adj} = D_{unadjusted} \alpha$
- calculate $N_{adj}$ as $D_{adj}A$
- do this for for every spatial unit using the adjustment that applies in that particular region (the BCR subunit that the region is nested inside)
- same applies to land cover classes within that BCR subunit
- note that this needs to be repeated for all impacted species independently

For maps:

- use $D_{adj} = D_{unadjusted} \alpha$ at the pixel level
- for this we can make a raster with adjustment values added for the cells matching the cell's BCR subunit designation
- take the adjustment layer and multiply with the density layer
- note that this needs to be repeated for all impacted species independently
